---
title: "fucus productivity"
author: "Lydia White & Leena Virta"
date: "2023-05-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

First we load libraries (which you should already have installed).

```{r}
library(tidyverse)
library(googlesheets4)
library(ggtext)
library(car)
```

## Visualise differences in NPP

Load data. 

```{r}
gs4_deauth()

productivity <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "fucus_productivity_summary") 

```

Visualise with boxplot of NPP and CR and GPP, per g of fucus for different sites.
We can make less work for ourselves and plot all the data at once, but we have to reformat the data into long format first.

```{r}

ggplot(productivity, aes(x=site, y= NPP, fill = site))+
  geom_boxplot()

productivity_long <- productivity %>% pivot_longer(cols = c("NPP", "CR", "GPP"), names_to = "metabolism", values_to = "carbon_flux") 
  
ggplot(productivity_long, aes(x=site, y= carbon_flux, fill = site))+
  geom_boxplot() +
  facet_wrap(~metabolism)

```

We can tidy up the plot a bit, we 1) edit the y axis label and 2) remove the legend 

```{r}

ggplot(productivity_long, aes(x=site, y= carbon_flux, fill = site))+
  geom_boxplot() +
  facet_wrap(~metabolism) +
  ylab("g carbon per g fucus per hour") +
  theme(legend.position="none")

```

## Test differences with models

Lets explore these differences with statistical models, we can use one-way ANOVA to test if carbon fluxes significantly differ between sites 

```{r}

anova1 <- aov(NPP ~ site, data = productivity)
summary(anova1)
```

### Checking assumptions of your model

The ANOVA test assumes that, the residuals from the model are normally distributed and the variance across groups are homogeneous (i.e. similar). We can check that with some diagnostic plots.

#### Homogeneity of variances

The residuals versus fits plot can be used to check the homogeneity of variances, the line. shoud be approximately straight

```{r}
plot(anova1, 1)

```

It’s also possible to use Levene’s test to check the homogeneity of variances. p < 0.05 means that they are not homogeneous. 

```{r}

leveneTest(NPP ~ site, data = productivity)
```


#### Normally distributed residuals

The normal probability plot of residuals is used to check the assumption 
that the residuals are normally distributed. 
It should approximately follow a straight line.

```{r}
plot(anova1, 2)
```

It’s also possible to use shapiro wilk test to check if the residuals are normally distributed. p < 0.05 means that they are not normally distributed.  

```{r}
shapiro.test(rstandard(anova1))

```


In one-way ANOVA test, a significant p-value indicates that some of the group means are different, but we don’t know which pairs of groups are different.
It’s possible to perform multiple pairwise-comparison, to determine if the mean difference between specific pairs of group are statistically significant.

```{r}
TukeyHSD(anova1)
```

### What if assumptions are not met?

Note that, a non-parametric alternative to one-way ANOVA is Kruskal-Wallis 
rank sum test, which can be used when ANOVA assumptions are not met.

```{r}
kruskal.test(NPP ~ site, data = productivity)

```

## Differences in CR and GPP between sites

You could also carry out ANOVAs on the other parameters, to see if CR and GPP differ between sites. 

```{r}
# write scripts here (copy and paste from above and change variable names)

```


