---
title: "Fauna community analysis"
author: "Lydia White, Leena Virta & Nico Henry"
date: "2023-05-18"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Set up our environment

### Load libraries

First we load libraries (which you should already have installed), we will also need vegan for doing multivariate analysis 

```{r}
library(tidyverse)
library(googlesheets4)
library(vegan)
```

## Abundance 

### Import data

```{r}
gs4_deauth()
abundance <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg" , sheet = "benthos_abundance_m2")
```

### Visualise differences in fauna abundance

Boxplot of total abundance per m2 for different sites 

```{r}
ggplot(abundance, aes(x=site, y= total_abundance, fill = site)) + 
  geom_boxplot()
```

We can tidy up the plot a bit, we edit the y axis label and remove the legend 

```{r}
ggplot(abundance, aes(x=site, y=total_abundance, fill=site)) + 
  geom_boxplot() +
  ylab("total macrofauna abundance per m2") +
  theme(legend.position="none") 

```

You can do the same for species richness, but we need to seperate the species abundance matrix and site information into 2 different dataframes. 

R doesn't recognise empty spaces from the excel file, we need to tell it they are zeros. 
So we replace NA with zero on all numeric columns. 

```{r}
species_matrix <- abundance %>% select (-site, -date, -replicate, -treatment, -total_abundance) %>% data.frame(row.names = "sample")

environmental_table <- abundance %>% select (sample, site, date, replicate, treatment, total_abundance) 

species_matrix <- mutate_all(species_matrix, ~as.numeric(replace_na(.,0)))

richness <- specnumber(species_matrix) 

richness <- tibble(sample = names(richness), species_richness = richness) %>% 
  inner_join( environmental_table)

ggplot(richness, aes(x=site, y=species_richness, fill=site)) + 
  geom_boxplot() +
  ylab("mean species richness per quadrat") +
  theme(legend.position="none") 


```

### Test differences with models

One way ANOVA 

```{r}
anova1 <- aov(species_richness~ site, data = richness)
summary(anova1)
```

Check assumptions etc... 

```{r}
# write script here (copy and paste from previous scripts and change variable names etc)
```


### Multivariate analysis 

Standardise from absolute abundance to relative abundance

```{r}

species_matrix_rel <-         
  decostand(species_matrix, method = "total")

```

#### NMDS plot

Calculate distance matrix and plot NMDS. 

```{r}
species_matrix_distmat <- 
  vegdist(species_matrix_rel, method = "bray")

species_matrix_NMDS <-
  metaMDS(species_matrix_distmat,
          distance = "bray",
          k = 2,
          maxit = 999, 
          trymax = 500,
          wascores = TRUE)

scores <- as_tibble(scores(species_matrix_NMDS), rownames = "sample")

final <- inner_join(environmental_table, scores)

ggplot(final, aes(x= NMDS1, y= NMDS2, colour = site)) + 
  geom_point( ) + 
  stat_ellipse()

```

#### PERMANOVA 

We can do a PERMANOVA, analogous to a ANOVA, but for multivariate data, to see if communities differ between site. We need to check the assumptions of homogeneous dispersion between groups before doing a PERMANOVA. p < 0.05 means that they are not homogeneous. 

```{r}
disper <- betadisper(species_matrix_distmat, environmental_table$site)
plot(disper)
anova(disper)
```

then we run the PERMANOVA.

```{r}
adonis2(species_matrix_rel ~ site, data = environmental_table, permutations = 999)

```

Then pairwise tests to see which groups differ. You can load a special function for this. See for more info: https://github.com/pmartinezarbizu/pairwiseAdonis

```{r}
source(here::here("R", "pairwise_adonis2.R"))

pairwise.adonis2 (species_matrix_rel ~ site, data = environmental_table, permutations = 999)
```

You can explore which taxa contribute the most to identified differences between groups using SIMPER. 

SIMPER is intended for comparisons among levels of a categorical variable, where the response matrix is expressed as a distance matrix, particularly the Bray-Curtis dissimilarity measure.

```{r}
test<- with(environmental_table, simper(species_matrix_rel, site))
summary(test)

```

## Biomass  

You could also do analysis on total biomass and multivariate analysis on biomass instead of abundance. You could copy and adapt the scripts for the biomass data. 

### Import data

```{r}
biomass <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "benthos_biomass_m2")
```

### Visualise and run models

```{r}

ggplot(biomass, aes(x=site, y=total_biomass, fill=site)) + 
  geom_boxplot() +
  ylab("g total macrofauna biomass per m2") +
  theme(legend.position="none") 

# write scripts here (copy and paste from above and change variable names etc)

```

### Multivariate analysis

```{r}
species_matrix <- biomass %>% select (-site, -date, -replicate, -treatment, -total_biomass, -total_biomass_dry, -total_biomass_carbon) %>% data.frame(row.names = "sample")

environmental_matrix <- biomass %>% select (sample, site, date, replicate, treatment, total_biomass)

species_matrix <- mutate_all(species_matrix, ~as.numeric(replace_na(.,0)))

# write scripts here (copy and paste from above and change variable names etc)
```


