---
title: "Sediment fluxes"
author: "Lydia White & Leena Virta"
date: "2023-05-22"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

First we load libraries (which you should already have installed). 

```{r}
library(tidyverse)
library(googlesheets4)
library(patchwork)
library(ggpubr)
```


# Sediment fluxes

We load the data. 

```{r}
gs4_deauth()
fluxes <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "sediment_fluxes") 

```

## Visualise differences in sediment fluxes

Visualise with boxplots of O2 and ammonium for different sites. 

```{r}
ggplot(fluxes, aes(x=site, y= `O2_flux µg per square metre per hour`, fill = site)) + 
  geom_boxplot() +
  ylab("µg O2 per square metre per hour") +
  theme(legend.position="none")

ggplot(fluxes, aes(x=site, y= `NH4_flux µg per square metre per hour`, fill = site)) + 
  geom_boxplot() +
  ylab("µg NH4-N per square metre per hour") +
  theme(legend.position="none")

```

You could also test for differences with models (ANOVA, Kruskal-Wallis etc)

```{r}
# write script here (copy and paste from previous scripts and change variable names etc)
```

## Are fluxes affected by fauna?

We can explore how fluxes from the sediment are affected by faunal communities via scatter plots. We also need to load in the fauna data. 

### Abundance & Biomass

```{r}
abundance <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "benthos_abundance_m2")

abundance <- abundance %>% select(sample, total_abundance)
  
biomass <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "benthos_biomass_m2")
biomass  <- biomass %>% select(sample, total_biomass)

```

We can join the flux data and the fauna together as a single dataframe, we join them by their unique sample ID 

```{r}
summary <- inner_join(fluxes, abundance, by = "sample") %>%
  inner_join(., biomass, by = "sample")

```

### Visualise relationshps with scatter plot 

```{r}

plot_1 <- ggplot(summary, aes(x=total_abundance, y= `O2_flux µg per square metre per hour`)) + 
  geom_point() 

plot_2 <- ggplot(summary, aes(x=total_biomass, y= `O2_flux µg per square metre per hour`)) + 
  geom_point() 

plot_3 <- ggplot(summary, aes(x=total_abundance, y= `NH4_flux µg per square metre per hour`)) + 
  geom_point() 

plot_4 <- ggplot(summary, aes(x=total_biomass, y= `NH4_flux µg per square metre per hour`)) + 
  geom_point() 

(plot_1 + plot_2)/
  (plot_3 + plot_4)

```


### Check assumptions for correlation

Check if data is normal for the 4 variables with shapiro-wilk test, p < 0.05 means data is not normally distributed. 
If data is normal use pearsons correlation, if its non-normal then use spearmans rank.

```{r}
shapiro.test(summary$total_abundance)
shapiro.test(summary$total_biomass)
shapiro.test(summary$`O2_flux µg per square metre per hour`)
shapiro.test(summary$`NH4_flux µg per square metre per hour`)
```

Some are normal, some might not be, we can replot the data showing significant correlations. 
Adjust which correlation test you use depending if variables are non-normal or normal. 
Adapt examples below for your data. 

```{r}
plot_1 <- ggscatter(summary, x = "total_abundance", y = "O2_flux µg per square metre per hour", 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "total fauna abundance", ylab = "oxygen flux")

plot_2 <- ggscatter(summary, x = "total_biomass", y = "O2_flux µg per square metre per hour", 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "total fauna biomass", ylab = "oxygen flux")

# adapt script above, using the correct correlation test, add script here for plots 3 and 4 and put together in one plot 

(plot_1 + plot_2)/
  (plot_3)


```

If one of them is significant we can redo figure with added significance line. 

```{r}

plot_3 <- ggscatter(summary, x = "total_abundance", y = "NH4_flux µg per square metre per hour", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "total fauna abundance", ylab = "ammonium flux")

(plot_1 + plot_2)/
  (plot_3 + plot_4)
```

## Are fluxes affected by organic matter in the sediment?


