---
title: "Sediment fluxes"
author: "Lydia White & Leena Virta"
date: "2023-05-22"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```

## Set up our environment

### Load libraries

First we load libraries (which you should already have installed).

```{r}
library(tidyverse)
library(googlesheets4)
library(patchwork)
library(ggpubr)
```

## Import data

```{r}
gs4_deauth()
fluxes <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "sediment_fluxes") 

```

## Visualise differences in sediment fluxes

Visualise with boxplots of O2 and ammonium for different sites. 

```{r}
ggplot(fluxes, aes(x=site, y= `O2_flux µg per square metre per hour`, fill = treatment)) + 
  geom_boxplot() +
  ylab("µg O2 per square metre per hour")

ggplot(fluxes, aes(x=site, y= `NH4_flux µg per square metre per hour`, fill =  treatment)) + 
  geom_boxplot() +
  ylab("µg NH4-N per square metre per hour")

```

You could also test for differences with models (ANOVA, Kruskal-Wallis etc), you could do this separately for light and dark treatments.

```{r}
light <- fluxes %>% filter (treatment =="light")
dark <- fluxes %>% filter (treatment =="dark")
  
# write script here (copy and paste from previous scripts and change variable names etc)
```

## Are fluxes affected by fauna?

We can explore how fluxes from the sediment are affected by faunal communities via scatter plots. We also need to load in the fauna data. 

### Import data
```{r}
abundance <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "benthos_abundance_m2")
abundance <- abundance %>% select(sample, total_abundance)
  
biomass <- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "benthos_biomass_m2")
biomass  <- biomass %>% select(sample, total_biomass)

```

We can join the flux data and the fauna together as a single dataframe, we join them by their unique sample ID. 
Lets explore the dark data, i.e. when organisms are respiring.

```{r}
summary <- inner_join(dark, abundance, by = "sample") %>%
  inner_join(., biomass, by = "sample")

```

### Visualise relationships with scatter plots 

```{r}

plot_1 <- ggscatter(summary, x = "total_abundance", y = "O2_flux µg per square metre per hour")
plot_2 <- ggscatter(summary, x = "total_abundance", y = "NH4_flux µg per square metre per hour")
plot_3 <- ggscatter(summary, x = "total_biomass", y = "O2_flux µg per square metre per hour")
plot_4 <- ggscatter(summary, x = "total_biomass", y = "NH4_flux µg per square metre per hour")

(plot_1 + plot_2)/
  (plot_3 + plot_4)

```

### Check assumptions for correlation

Check if data is normal for the 4 variables with shapiro-wilk test, p < 0.05 means data is not normally distributed. 
If data is normal use pearsons correlation, if its non-normal then use spearmans rank.

```{r}
shapiro.test(summary$total_abundance)
shapiro.test(summary$total_biomass)
shapiro.test(summary$`O2_flux µg per square metre per hour`)
shapiro.test(summary$`NH4_flux µg per square metre per hour`)
```

Some are normal, some might not be, we can replot the data showing correlation coefficents. 
Adjust which correlation test you use depending if variables are non-normal or normal. 
Adapt examples below for your data. 

```{r}
plot_1 <- ggscatter(summary, x = "total_abundance", y = "O2_flux µg per square metre per hour",
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "total fauna abundance", ylab = "oxygen flux") 


plot_2 <- ggscatter(summary, x = "total_abundance", y = "NH4_flux µg per square metre per hour",
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "total fauna abundance", ylab = "ammonium flux") 

# adapt script above, using the correct correlation test, add script here for plots 3 and 4 and put together in one plot 

(plot_1 + plot_2)/
  (plot_3 + plot_4)


```

If one of them is significant we can redo figure with added significance line. Adapt to whichever plot is significant. 

```{r}
plot_1 <- ggscatter(summary, x = "total_abundance", y = "O2_flux µg per square metre per hour",
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson",
          xlab = "total fauna abundance", ylab = "oxygen flux") 

# adapt script to add line for whichever plots have significant variables. 

(plot_1 + plot_2)/
  (plot_3 + plot_4)
```

## Are fluxes affected by organic matter in the sediment?

We can also explore how fluxes from the sediment are affected by the amount of carbon in the sediment via scatter plots. We also need to load in the carbon stocks data. 

### Import data

```{r}
carbon_stocks<- read_sheet("1_HAbVc0HTqO6iMY_jTmrq6zVtyEmIokK3xqg6-RjNVg", sheet = "carbon_stocks_m2")

sed_summary <- inner_join(dark, carbon_stocks, by = c("site", "replicate"))
```

### Visualise relationships with scatter plots 

```{r}

plot_5 <- ggscatter(sed_summary, x = "C_sediment", y = "O2_flux µg per square metre per hour")
plot_6 <- ggscatter(sed_summary, x = "C_sediment", y = "NH4_flux µg per square metre per hour")

plot_5 + plot_6
```


### Check assumptions for correlation

Check if data is normal for the sediment carbon content with shapiro-wilk test, p < 0.05 means data is not normally distributed. 
If data is normal use pearsons correlation, if its non-normal then use spearmans rank.

```{r}
shapiro.test(sed_summary$C_sediment)

```

We can replot with correlation coefficients. 

```{r}

plot_5 <- ggscatter(sed_summary, x = "C_sediment", y = "O2_flux µg per square metre per hour",
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "sediment carbon content", ylab = "oxygen flux") 

plot_6 <- ggscatter(sed_summary, x = "C_sediment", y = "NH4_flux µg per square metre per hour",
          cor.coef = TRUE, cor.method = "spearman",
          xlab = "sediment carbon content", ylab = "ammonium flux") 

(plot_1 + plot_2) /
  (plot_3 + plot_4) /
    (plot_5 + plot_6 )

```





